---
title: "Homework 4"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)
options(width = 100)
```



```{r install-packages, eval=FALSE}
install.packages(c("DBI", "RMariaDB", "mdsr", "knitr", "rmarkdown"))
```

```{r load-packages, warning=FALSE}
library(DBI)
library(RMariaDB)
library(mdsr)
```

## Q2

```{r q2}
con_wai <- dbConnect(
  MariaDB(),
  host = "scidb.smith.edu",
  user = "waiuser",
  password = "smith_waiDB",
  dbname = "wai"
)

q2 <- dbGetQuery(con_wai, "
  SELECT COUNT(*) AS n_rows
  FROM Measurements;
")
knitr::kable(q2)
dbDisconnect(con_wai)
```

The `Measurements` table contains **`r format(as.numeric(q2$n_rows), big.mark = ",", trim = TRUE)` rows**.

## Q3

```{r q3}
con_air <- dbConnect_scidb("airlines")

q3 <- dbGetQuery(con_air, "
  SELECT DISTINCT year
  FROM flights
  ORDER BY year;
")
knitr::kable(q3)
```

The available years are **`r paste(q3$year, collapse = ", ")`**.
`DISTINCT` returns each year only once.

## Q4.1

```{r q4-1}
q4_1 <- dbGetQuery(con_air, "
  SELECT COUNT(*) AS n_flights,
         SUM(cancelled = 1) AS n_cancelled,
         SUM(cancelled = 0 AND diverted = 0) AS n_completed
  FROM flights
  WHERE year = 2015 AND month = 5 AND day = 14
    AND dest = 'DFW';
")
knitr::kable(q4_1)
```

There are **`r q4_1$n_flights` flight records** with DFW as the destination on
May 14, 2015. This total includes `r q4_1$n_cancelled` cancelled flights.
If "flew into" means non-cancelled, non-diverted arrivals, the answer is
**`r q4_1$n_completed` flights**.

## Q4.2

```{r q4-2}
q4_2 <- dbGetQuery(con_air, "
  SELECT dest, COUNT(*) AS n_flights
  FROM flights
  WHERE year = 2015 AND origin = 'ORD'
  GROUP BY dest
  ORDER BY n_flights DESC, dest
  LIMIT 6;
")
knitr::kable(q4_2, format.args = list(big.mark = ","))
```

The six most common destinations were **`r paste(q4_2$dest, collapse = ", ")`**,
in descending order. **`r q4_2$dest[1]`** was the most common, with
**`r format(as.numeric(q4_2$n_flights[1]), big.mark = ",", trim = TRUE)` flights**.

## Q4.3

```{r q4-3}
q4_3 <- dbGetQuery(con_air, "
  SELECT dest,
         AVG(arr_delay) AS avg_arr_delay_minutes,
         COUNT(arr_delay) AS n_observed_delays
  FROM flights
  WHERE year = 2015
  GROUP BY dest
  ORDER BY avg_arr_delay_minutes DESC, dest
  LIMIT 1;
")
knitr::kable(q4_3, digits = 2)
```

**`r q4_3$dest`** had the highest average arrival delay, approximately
**`r sprintf("%.2f", q4_3$avg_arr_delay_minutes)` minutes**, based on
`r q4_3$n_observed_delays` non-missing delay values.
`AVG()` excludes SQL `NULL` values. Negative delays, representing early arrivals,
are retained, and no minimum flight-count threshold is imposed.

## Q4.4

```{r q4-4}
q4_4 <- dbGetQuery(con_air, "
  SELECT COUNT(*) AS n_flights,
         SUM(dest = 'BDL') AS n_inbound,
         SUM(origin = 'BDL') AS n_outbound,
         SUM(cancelled = 1) AS n_cancelled,
         SUM(
           cancelled = 0 AND
           (origin = 'BDL' OR (dest = 'BDL' AND diverted = 0))
         ) AS n_operated_to_or_from_bdl
  FROM flights
  WHERE year = 2015
    AND (origin = 'BDL' OR dest = 'BDL');
")
knitr::kable(q4_4, format.args = list(big.mark = ","))
```

There are **`r format(as.numeric(q4_4$n_flights), big.mark = ",", trim = TRUE)` flight records**
to or from BDL in 2015: `r format(q4_4$n_inbound, big.mark = ",", trim = TRUE)`
inbound and `r format(q4_4$n_outbound, big.mark = ",", trim = TRUE)` outbound.
The parentheses ensure that the year condition applies to both directions.

The total includes `r q4_4$n_cancelled` cancelled flights.
Excluding cancellations and inbound flights diverted away from BDL gives
**`r format(q4_4$n_operated_to_or_from_bdl, big.mark = ",", trim = TRUE)` flights**
that operated to or from BDL according to these fields. A non-cancelled flight
departing BDL is included even if it was later diverted.

## Q4.5

```{r q4-5}
q4_5 <- dbGetQuery(con_air, "
  SELECT c.name AS airline, f.carrier, f.flight,
         f.origin, f.dest
  FROM flights AS f
  LEFT JOIN carriers AS c ON f.carrier = c.carrier
  WHERE f.year = 2015 AND f.month = 9 AND f.day = 26
    AND (
      (f.origin = 'LAX' AND f.dest = 'JFK')
      OR (f.origin = 'JFK' AND f.dest = 'LAX')
    )
  ORDER BY f.origin, c.name, f.flight;
")
knitr::kable(q4_5)
```

The table lists all flights in **both directions**. 
```{r disconnect, include=FALSE}
dbDisconnect(con_air)
```
